---
title: "Lab 3 - Survival Analysis"
author: "Richard CHEAM"
date: "2024-10-17"
output: html_document
---

# Import libraries

```{r}
library(KMsurv) #data
library(tidyverse) #operator %>%
library(survival) #tmerge()
library(ggplot2)
library(ggfortify)
```

# Load dataset

```{r}
data(bmt)
```

```{r}
head(bmt)
```

```{r}
cat("Dimension of dataset:", dim(bmt))
```

Covariates are $z1, z2, ..., z10$. 
T is $t2$ and indicator sigma is $d3$.
Time-dependent variable is of $tp$ and $dp$.

# Question 1

```{r}
transform <- function(df){
  #rename 
  new_df <- df %>%
  rename(
    'agep' = 'z1',
    'aged' = 'z2',
    'genderp' = 'z3',
    'genderd' = 'z4',
    'cmvp' = 'z5',
    'cmvd' = 'z6',
    'waiting' = 'z7',
    'FAB' = 'z8',
    'hospital' = 'z9',
    'MTW' = 'z10',
    'DFS' = 't2',
    'DFSstatus' = 'd3'
  )
  #recode
  new_df$group <- replace(new_df$group, new_df$group == 1, 'ALL')
  new_df$group <- replace(new_df$group, new_df$group == 2, 'Low')
  new_df$group <- replace(new_df$group, new_df$group == 3, 'High')
  #centered age
  new_df$agep <- replace(new_df$agep, new_df$agep, new_df$agep - mean(new_df$agep))
  new_df$aged <- replace(new_df$aged, new_df$aged, new_df$agep - mean(new_df$aged))
  #remove columns
  new_df <- subset(new_df, select = -c(t1, d1, d2, ta, da, tc, dc))
  return(new_df)
}
```


Rename variables:

```{r}
bmt <- bmt %>%
  rename(
    'agep' = 'z1',
    'aged' = 'z2',
    'genderp' = 'z3',
    'genderd' = 'z4',
    'cmvp' = 'z5',
    'cmvd' = 'z6',
    'waiting' = 'z7',
    'FAB' = 'z8',
    'hospital' = 'z9',
    'MTW' = 'z10',
    'DFS' = 't2',
    'DFSstatus' = 'd3'
  )
```

```{r}
colnames(bmt)
```


Recode values taken by $group$

```{r}
unique(bmt['group'])
```

```{r}
bmt$group <- replace(bmt$group, bmt$group == 1, 'ALL')
bmt$group <- replace(bmt$group, bmt$group == 2, 'Low')
bmt$group <- replace(bmt$group, bmt$group == 3, 'High')
```

```{r}
unique(bmt$group)
```

```{r}
mean(bmt$agep)
mean(bmt$aged)
```

```{r}
bmt$agep <- replace(bmt$agep, bmt$agep, bmt$agep - mean(bmt$agep))
bmt$aged <- replace(bmt$aged, bmt$aged, bmt$agep - mean(bmt$aged))
```

```{r}
bmt$aged
```

```{r}
bmt2 <- subset(bmt, select = -c(t1, d1, d2, ta, da, tc, dc))
```

New dataset:

```{r}
head(bmt2)
```

```{r}
cat("Dimension of dataset:", dim(bmt2))
```

```{r}
colnames(bmt2)
```

# 2. Create time-varying variable

Add id column to dataframe

```{r}
bmt2$id <- seq(1, nrow(bmt2))
```

```{r}
bmt2
```


$DFS$ is T and $DFSstatus$ is indicator sigma. 

```{r}
bmt2_merge <- tmerge(bmt2, bmt2, id = id, event = event(DFS, DFSstatus))
```

```{r}
bmt2_merge
```

```{r}
bmt2_merge <- tmerge(bmt2_merge, bmt2, id = id, platelet_recov = tdc(tp))
```

```{r}
bmt2_merge$platelet_recov
```

Each value in tp corresponds to the time when platelet recovery occurs for an individual. The tdc(tp) will allow you to model this as a time-dependent covariate that changes its value when the time reaches the specified value for each individual. The tdc(tp) function will dynamically update the covariate platelet_recov at the times specified in tp for each individual. Before tp, the covariate is 0 (no recovery), and after tp, the covariate changes (indicating recovery).

# 3. Construct appropriate cox model

## Factor categorical variables

```{r}
bmt2_merge$FAB <- replace(bmt2_merge$FAB, bmt2_merge$FAB == 1, 'FAB Grade 4 Or 5 and AML')
bmt2_merge$FAB <- replace(bmt2_merge$FAB, bmt2_merge$FAB == 0, 'Otherwise')

bmt2_merge$hospital <- replace(bmt2_merge$hospital, bmt2_merge$hospital == 1, 'The Ohio State University')
bmt2_merge$hospital <- replace(bmt2_merge$hospital, bmt2_merge$hospital == 2, 'Alferd')
bmt2_merge$hospital <- replace(bmt2_merge$hospital, bmt2_merge$hospital == 3, 'St.Vincent')
bmt2_merge$hospital <- replace(bmt2_merge$hospital, bmt2_merge$hospital == 4, 'Hahnemann')

bmt2_merge$MTW <- replace(bmt2_merge$MTW, bmt2_merge$MTW == 1, 'Yes')
bmt2_merge$MTW <- replace(bmt2_merge$MTW, bmt2_merge$MTW == 0, 'No')
```

```{r}
bmt2_merge
```



## Remove unwanted 

```{r}
bmt2_merge_for_fit <- subset(bmt2_merge, select = -c(DFS, DFSstatus, dp, tp, id))
bmt2_merge_for_fit
```

```{r}
colnames(bmt2_merge_for_fit)
```


## Fit cox model

```{r}
cox_model <- coxph(formula = Surv(tstart, tstop, event) ~., data = bmt2_merge_for_fit)
```

```{r}
summary(cox_model)
```

## Variables selection

Selected variables are: 

* z8 (FAB): 1-FAB Grade 4 Or 5 and AML, 0-Otherwise (The French, American, and British (FAB) classification system for acute myeloid leukemia (AML))
* z9 (hospital): 1-The Ohio State University, 2-Alferd , 3-St. Vincent, 4-Hahnemann
* z10 (MTW): MTX Used as a Graft-Versus-Host- Prophylactic: 1-Yes 0-No

```{r}
cox_model_selected_vars <- coxph(formula = Surv(tstart, tstop, event) ~ FAB + hospital + MTW, data = bmt2_merge_for_fit)
summary(cox_model_selected_vars)
```

The p-value for all three overall tests are significant, indicating that the model is significant. These tests give the confidence to reject null hypothesis that all of the betas are 0.

The 3 covariates remain significant as p-values are below the threshold 0.05 and have positive impact on survival.

The p-value of FAB is 0.000973 with a hazard ratio (exp(coef)) of 2.0949 indicating a strong relationship between the patients’ FAB classification grade and disease free survival time (Dead/Relapsed or Alive/100% Recovered). FAB increases the hazard by factor of 2.0949.

While hospital reduces the hazard by factor of 0.6828

## Plot

```{r}
autoplot(survfit(cox_model_selected_vars), color = "#2E9FDF",
           ggtheme = theme_minimal())
```

We want to see effect of each covariate:

In this case, we construct a new data frame with two rows, one for each value of FAB; the other covariates are fixed to their lowest level since they are discrete variables. 

