---
title: "TP2"
author: "Richard CHEAM"
date: "2024-10-03"
output: html_document
---

```{r}
library(survival)
library(KMsurv)
library(ggplot2)
library(ggfortify)
```

# Load data

```{r}
data(pneumon)
pneumon
```

```{r}
cat("Dimension of dataframe:", dim(pneumon))
```

# 1. Check if the variables have been correcly imported, especially the factors

```{r}
colnames(pneumon)
```

```{r}
unique(pneumon$agepn)
```

```{r}
unique(pneumon$chldage)
```

```{r}
pneumon <- subset(pneumon, select = -agepn)
pneumon
```

```{r}
table(pneumon$hospital)
```


# 2. Plot the Kaplan-Meier estimator for the survival function of the age at pneumonia. Give an estimation and a confidence interval for the probability for a newborn of not having developed pneumonia at 6 months.

```{r}
KM_fit <- survfit(Surv(chldage, hospital) ~ 1, data = pneumon)
summary(KM_fit)
```

```{r}
autoplot(KM_fit, xlab="Time (months)", ylab="Survival Probability")
```
From the summary table: the probability of a newborn not having developed pneumonia is 0.981 with 95% interval [0.976, 0.986].

# Construct a dummy variable Z = 1 if infants were breast fed at birth, 0 if infants were never breast fed, and test the hypothesis H0: there is no difference in distributions of age at first pneumonia between a child with breast fed and one without.

Z = 1 (breast fed), 0 (not)

```{r}
Z <- numeric(dim(pneumon)[1])
for (i in 1:dim(pneumon)[1]){
  if (pneumon$wmonth[i] > 0){
    Z[i] = 1
  }
  else {
    Z[i] = 0
  }
}
```

Add column Z to DataFrame pneumon.

```{r}
pneumon$breastFed <- Z
```

```{r}
pneumon
```

```{r}
#object = Surv(chldage, hospital) ~ 1, data = pneumon
#to fit = use survfit
KM_fit <- survfit(Surv(chldage, hospital) ~ breastFed, data = pneumon)
```

```{r}
summary(KM_fit)
```
```{r}
autoplot(KM_fit, xlab = "Time (months)", ylab = "Survival Probability")
```

# 4. Test the hypothesis H0 : βbreastf = 0, i.e., the survival functions for the two types of breast feeding are equal, using the likelihood ratio, and Wald tests. Find the estimate of βbreastf ⋆ , β^breastf ⋆ , its standard error, and the relative risk

```{r}
cox_fit <- coxph(Surv(chldage, hospital) ~ breastFed, data = pneumon)
```

```{r}
summary(cox_fit)
```
p-value of these tests are really which mean we can reject H0 which states that breastFed does not affect. Hence, breastFed has effect.

Estimated coefficient is -1.0970.
Relative risk is 0.3339. 
Standard error is 0.2973.

```{r}
summary(cox_fit)$logtest
summary(cox_fit)$waldtest
```

# 5. Test the hypothesis that the times to hospitalized pneumonia are the same for the two feeding groups adjusting for each of these factors in a separate model using the Wald test.

Iterate through all column as an additional to breastFed model above. Starting from mtage to sfmonth.

```{r}
colnames(pneumon)
```


```{r}
for (var in colnames(pneumon)[3:14]){
  print(var)
  fit = coxph(as.formula(paste("Surv(chldage, hospital) ~ breastFed + ", var)), data = pneumon)
  print(summary(fit))
  print("++++++++++++++++++++++++++++++++++++++++++++++++")
}
```

# 6. Build a model with the view of testing the particular comparison of interest adjusting for the other non controllable fixed covariates in question 4

```{r}
summary(coxph(Surv(chldage, hospital) ~ ., data = pneumon))
```

# 7. In the final model, predict the probability of not developing pneumonia at 7 months for a newborn with covariates

```{r}
summary(survfit(Surv(chldage, hospital) ~ 1, data = pneumon))
```


