---
title: "tp1-sur"
author: "Richard CHEAM"
date: "2024-09-26"
output: pdf_document
---
## Import libraries

```{r}
library(MASS)
library(asaur)
library(survival)
library(KMsurv)
library(ggplot2)
library(ggfortify)
```

## Read dataframe

```{r}
df <- read.csv(file = "coalition2.csv", header = TRUE)
head(df)
```

```{r}
cat("Dimension of the dataframe:", dim(df))
```

### 1. Compute and draw the empirical c.d.f. for the variable duration.

```{r}
X = df$duration

emp_cdf = ecdf(X)
plot(emp_cdf, col = 'red', main = 'Empirical CDF', xlab = 'duration')
```

```{r}
ggplot(data = data.frame(duration = X), aes(x = duration)) + 
  stat_ecdf(geom = "step", color = "blue", size = 1) +  
  theme_minimal() + 
  labs(
    title = "Empirical Cumulative Distribution Function (CDF) Plot",
    x = "Duration",
    y = "Cumulative Probability"
  )
```


### 2. Compute maximum likelihood estimator for the variable duration for 4 different families.

Parameters of each family are shown below:

```{r}
fit_normal <- fitdistr(X, "normal")
fit_weibull <- fitdistr(X, "weibull")
fit_lognormal <- fitdistr(X, "lognormal")
fit_gamma <- fitdistr(X, "gamma")
fit_normal
fit_weibull
fit_lognormal
fit_gamma
```

### 3. Draw the estimated c.d.f. in each family on the same plot as the empirical c.d.f

```{r}
cdf_normal <- pnorm(X, mean = fit_normal$estimate['mean'], sd = fit_normal$estimate['sd'])

cdf_weibull <- pweibull(X, shape = fit_weibull$estimate['shape'], scale = fit_weibull$estimate['scale'])

cdf_lognormal <- plnorm(X, meanlog = fit_lognormal$estimate['meanlog'], sdlog = fit_lognormal$estimate['sdlog'])

cdf_gamma <- pgamma(X, shape = fit_gamma$estimate['shape'], rate = fit_gamma$estimate['rate'])
```

```{r}
# Create a data frame for plotting
cdf_data <- data.frame(
  x = X,
  normal = cdf_normal,
  weibull = cdf_weibull,
  lognormal = cdf_lognormal,
  gamma = cdf_gamma
)

# Plot the empirical CDF and estimated CDFs
ggplot() +
  stat_ecdf(data = df, aes(x = duration), color = 'black', size = 1, linetype = "dashed") + 
  geom_line(data = cdf_data, aes(x = x, y = normal, color = 'Normal'), size = 1) +
  geom_line(data = cdf_data, aes(x = x, y = weibull, color = 'Weibull'), size = 1) +
  geom_line(data = cdf_data, aes(x = x, y = lognormal, color = 'Lognormal'), size = 1) +
  geom_line(data = cdf_data, aes(x = x, y = gamma, color = 'Gamma'), size = 1) +
  labs(x = "Duration", y = "Cumulative Probability", title = "Empirical and Estimated CDFs") +
  scale_color_manual(values = c('Empirical' = 'black', 'Normal' = 'blue', 'Weibull' = 'green', 'Lognormal' = 'red', 'Gamma' = 'purple')) +
  theme_minimal()
```

```{r}
# Plot the empirical CDF
plot(emp_cdf, main = "Empirical and Estimated CDFs", xlab = "Duration", ylab = "CDF", col = "black", lwd = 2, verticals = FALSE)

# Add the estimated CDFs using lines()
plot(X, cdf_normal, col = "blue")    # Normal
plot(X, cdf_weibull, col = "green")  # Weibull
plot(X, cdf_lognormal, col = "red")  # Log-normal
plot(X, cdf_gamma, col = "purple")   # Gamma

# Add a legend
legend("bottomright", legend = c("Empirical", "Normal", "Weibull", "Log-normal", "Gamma"),
       col = c("black", "blue", "green", "red", "purple"), lwd = 2, lty = c(1, 2, 3, 4, 5))
```

# 2. Your own code for the Kaplan-Meier estimator

## 1. Develop a function to compute the Kaplan-Meier estimator that takes as inputs

```{r}
my_KM <- function(time, sigma){
  t = sort(unique(time))
  return()
}

```

## 2. Consider the pharmocoSmoking data (available in package asaur), compare the results of your code to the one of the function survfit of package survival.

```{r}
data("pharmacoSmoking")
df <- pharmacoSmoking
head(df)
```

```{r}
dim(df)
```

```{r}
KM_fit <- survfit(Surv(df$ttr, df$relapse) ~ 1)
summary(KM_fit)
autoplot(KM_fit, xlab="Time", ylab="Survival Probability")
```

```{r}
print(KM_fit$surv)
cat("\n")
print(KM_fit$time)
```

```{r}
my_KM(df$ttr, df$relapse)
```


## 3. Compute the Greenwood estimator of the variance of the Kaplan-Meier estimator

```{r}
greenwood_est <- function(KM_est, nb_events, nb_at_risk, time){
  num_ = nb_events
  den_ = nb_at_risk * (nb_at_risk - nb_events)
  tmp = sum(num_/den_ < unique(time))
  return (KM_est**2 * tmp)
}

greenwood_est(KM_fit$surv, KM_fit$n.event, KM_fit$n.risk, KM_fit$time)
```

# Ex3. Left-truncated and right-censored data

```{r}
data("channing")
df <- channing
head(df)
```

```{r}
dim(df)
```

At age 901 how many residents are under observation and still alive
